\chapter{Related Work}
According to \cite{DBLP:journals/tist/ZhaoCYLDCWYD24}, the training of \acrshort{llm}s can be categorized into two paradigms, namely \textit{traditional fine-tuning} and \textit{prompting} depending on how they are adapted for downstream tasks (see \autoref{fig:llm_explainability}).
\\\\
In \textit{traditional fine-tuning}, a \acrlong{llm} is pre-trained on a large corpus of unlabeled text data and further fine-tuned on labeled data tailored to a specific domain and benchmarked accordingly (for example, the GLUE benchmark \cite{DBLP:conf/iclr/WangSMHLB19}). In the \textit{prompting} paradigm the so-called \textit{prompts} enable the model to behave in a certain way without retraining it, enabling zero-shot, one-shot and few-shot learning without requiring additional training data \cite{DBLP:conf/nips/BrownMRSKDNSSAA20}.
\\\\
As stated in \autoref{sec:xai} and shown in \autoref{fig:llm_explainability}, \textit{Gradient-Based Explanation} is a subcategory of \textit{feature attribution-based explanation}. However, the approach presented in this work falls under \textit{example-based explanation} (or in other words \textit{instance-based explanation}) but also uses a kind of gradient-based solution and therefore \textit{gradient-based explanation} will be briefly described in the next section.

\section{Gradient-Based Explanation}
The importance of each input feature can be evaluated using gradient-based attribution techniques. This is achieved by analyzing the partial derivatives of the output with respect to each input dimension. The magnitude of the derivatives reveals how sensitive the output is to changes in the input. This method can generally be formulated as $s_j=\frac{\partial f(x)}{\partial x_j}$, where $f(x)$ is the function of the neural network and $x_j$ is the input vector \cite{DBLP:journals/tist/ZhaoCYLDCWYD24}.
\\\\
However, the approach covered in this work will use the full gradients of each model layer with respect to a single training example instead of a single feature dimension; hence, it is more important to shed more light on example-based explanations and, therefore, different example-based variants will be briefly discussed in more detail below. Not all of them are actually completely related to the proposed approach in this work; however, it still makes sense to establish an outline of the closely and not so closely related topics to gain an overview before jumping into details of the proposed approach.
\\\\
Nevertheless, not all gradient-based explanation methods are tied to different feature dimensions, and gradients can also be useful in the context of example-based explanation. More about that is outlined in \autoref{subsec:data_influence}.

\section{Example-Based Explanation}
Example-based explanations offer a way to explain a model's behavior based on individual examples (or instances) \cite{DBLP:conf/icml/KohL17}. In other words, they offer a way to identify how a model's output changes on the basis of different inputs. Furthermore, according to \cite{DBLP:journals/tist/ZhaoCYLDCWYD24}, example-based explanations can be divided into three categories: \textit{adversarial examples}, \textit{counterfactual explanations}, and \textit{data influence}.

\subsection{Adversarial Example}
\acrlong{lm}s have been shown to be vulnerable to small adjustments in their input data. Although these carefully constructed changes may not be noticeable to humans, they can influence the output of a model. In the beginning, adversarial examples were based on word-level alterations such as small errors, removal, and insertion. Over time, more advanced token-level perturbation methods such as \textit{TextFooler}~\cite{DBLP:conf/aaai/JinJZS20} were published. These approaches specify important words on embedding similarity, same part of speech, etc. and are correspondingly modified. However, methods relying on word embeddings are limited in sentence representation compared to contextualized representations, which often lead to incoherent pieces.
\cite{DBLP:conf/naacl/WangXLCL22} proposed \textit{SemAttack}, a framework that can be used for different embedding spaces, such as type space, knowledge space, and contextualized semantic space. Once the input tokens are transformed to an embedding space, the perturbed embeddings are optimized iteratively until a specific goal is met. One of the experiments shows that by replacing 5\% of the words, the accuracy of BERT drops from 70.6\% to 2.4\%.

\subsection{Data Influence} \label{subsec:data_influence}
The notion of \textit{data influence} originally came from statistics. There, the influence of a particular data point on a model's parameters was measured by simply removing it during training. This concept, often called \textit{leave-one-out influence}, is a dominant framework for instance-based explanation. However, since retraining a model for every single data point is computationally infeasible, this effect is often approximated with re-training free methods. Regarding data influence methods, there are two important distinctions based on when gradients are obtained:
\subsubsection{Static, Gradient-Based Influence Estimation}
This type of explanation method got its name because it measures the influence only on the final model parameters $\theta_{T}$, which are static after training is finished. Compared to dynamic methods (see \nameref{subsubsec:dynamic_gradient_based_influence_estimation} below), the static estimator is computationally more efficient since it only relies on the final model parameters. However, it has less information to estimate and therefore provides limited insight. The approach proposed in this work also falls into this category, as it depends on an already trained or fine-tuned model and calculates gradients with respect to single input examples \cite{DBLP:journals/ml/HammoudehL24}.
\\\\
Using \textbf{influence functions}, a static data influence estimator, the authors of \cite{DBLP:conf/icml/KohL17} are able to explain the predictions of a black-box model by tracing back to the specific training examples that contributed the most to each prediction. To make this setting suitable for large-scale deep learning models, their approach relies on approximating these influence functions by using only gradient and Hessian-vector products.
\\\\
The contributors of \cite{DBLP:conf/emnlp/GuoRHBX21} demonstrate an improved application of influence functions for model debugging by using \acrfull{knn} to narrow down the search space to a subset of good candidate data points. However, computing a high-dimensional inverse Hessian-vector product is very expensive and, therefore, influence functions are difficult to scale to \acrlong{llm}s. Recent applications therefore restrict computation to a subset of model parameters, such as the multilayer perceptron (\acrshort{mlp}) layers \cite{DBLP:journals/corr/abs-2308-03296} or \acrfull{lora} layers \cite{kwon_datainf_2023}.
\\\\
Consequently, in 2023, Anthropic scaled the idea of influence functions for \acrshort{llm}s with up to 52 billion parameters by an \acrfull{ekfac} approximation. In their work, they use influence functions to examine generalization patterns of \acrlong{llm}s and evaluate their approximation by reporting agreement with the \textit{proximal Bergmann response function}. According to the authors, influences approach near zero when the order of the key phrases in a prompt is changed \cite{DBLP:journals/corr/abs-2308-03296}. Other evaluation strategies require expensive additional fine-tuning steps. For example, \cite{DBLP:conf/emnlp/AkyurekBLXTAG22} evaluate by introducing facts with the fine-tuning data that the model did not know before and then verify whether influence estimation methods correctly attribute importance to this fine-tuning set when queried about them. Such methods are often infeasible to perform with \acrlong{llm}s.

\subsubsection{Dynamic, Gradient-Based Influence Estimation} \label{subsubsec:dynamic_gradient_based_influence_estimation}
Dynamic gradient-based estimators approach the analysis by reconstructing the training's data influence based on the model parameters during training, e.g. $\theta_0,...,\theta_T$. An example of such an approach is TracInCP \cite{DBLP:conf/nips/PruthiLKS20}, which aims to approximate training data influence by comparing a test instance's loss gradient to the gradients of training instances at a set of model checkpoints. In the case of \acrlong{llm}s, working with the model parameters for each training checkpoint is more computationally expensive. However, they can be more precise than static estimators because they can harness more underlying information about the changes in the model with respect to the training data \cite{DBLP:journals/ml/HammoudehL24}.

\section{Paraphrased Corpora Analysis}
The authors of \cite{DBLP:conf/iclr/WangMS025} did an experiment that is similar to the approach proposed in this work, providing a re-training free evaluation setup. There, the robustness of In-Run Data Shapley and influence functions is evaluated by identifying relevant individual corpora that have been paraphrased. They also calculate a rank based on the original training corpus and its paraphrased versions. For comparison, the BM25~\cite{DBLP:journals/ftir/RobertsonZ09} distance is used as an oracle to assess the lexical similarity between the paraphrased versions and the original corpora, rather than the semantic similarity. Even with completely rewritten corpora, the corresponding original samples still rank very high according to In-Run Data Shapley and influence functions compared to BM25. This method allows for a systematic comparison of different strategies for obtaining instance-based explanations without costly re-training.

\section{Dimensionality and Performance}
The vast number of parameters in a modern \acrshort{llm}, often in the billions, makes analyzing the full gradient computationally prohibitive and potentially noisy. Therefore, selecting a subset of parameters is a common and necessary strategy to make instance-based gradient explanations feasible. This selection can be approached through traditional dimensionality reduction techniques or, more strategically, by focusing on specific, functionally significant components of the model's architecture, such as individual layers.

\subsection{Traditional Dimensionality Reduction Techniques}
One approach to manage the high dimensionality of gradients is to use methods like \textit{random projection} \cite{johnson_lindenstrauss}. This technique projects the high-dimensional gradient vector into a lower-dimensional space while approximately preserving the distances between vectors. While computationally efficient, random projection is data-agnostic and may obscure the fine-grained information encoded in the original gradient, as it does not consider the underlying structure of the neural network.

\subsection{Layer Selection}
A more informed approach to dimensionality reduction leverages the inherent structure of the model itself. Modern \acrshort{llm}s are predominantly built upon the Transformer architecture \cite{DBLP:conf/nips/VaswaniSPUJGKP17}, which is composed of a stack of identical layers. Each layer, or block, transforms its input representations through a multi-head self-attention mechanism and a position-wise feed-forward network. This deep, layered structure is not monolithic; instead, a functional specialization emerges across the network's depth.
\\\\
Research has shown that different layers capture different levels of linguistic abstraction, progressing from surface-level features to complex semantic representations \cite{DBLP:conf/acl/TenneyDP19, DBLP:conf/acl/JawaharSS19}. Generally, the \textit{initial layers} (closer to the input) tend to learn surface properties like word order and basic syntax. The \textit{middle layers} capture more complex syntactic and local semantic relationships, while the \textit{final layers} (closer to the output) synthesize this information into highly contextualized, task-specific representations.
\\\\
This hierarchical processing has significant implications for gradient-based explanation methods. The choice of which layer's parameters to use for gradient computation is not trivial, as gradients with respect to different layers can reveal which level of abstraction, syntactic, semantic, or otherwise, is most critical for a model's specific prediction. This raises a key question for instance-based explainability: which layer provides the most stable, informative, and computationally efficient signal for understanding model behavior?

\subsubsection{Are Early Layers Better?}
The common practice of restricting influence analysis to the final layer \cite{DBLP:conf/aaai/SchioppaZVS22, DBLP:conf/emnlp/AkyurekBLXTAG22} or just \acrshort{mlp} layers \cite{DBLP:journals/corr/abs-2308-03296} of a network, often chosen for computational convenience, has been challenged by recent work. The authors of \cite{DBLP:conf/nips/YehTSLR22} argue that for \acrshort{llm}s, the final layer is a suboptimal choice for calculating data influence due to a phenomenon they call \textit{cancellation effect}. This effect occurs when influence scores derived from later layers exhibit large magnitudes but conflict with one another, thereby reducing the discriminative power of the explanation method. The authors attribute this to the abstract, high-level features encoded in the network's deeper layers. In these layers, distinct inputs often activate the same neurons or pathways corresponding to generalized concepts, causing their gradients to become more aligned and, therefore, more susceptible to cancellation.
\\\\
Consequently, they propose that influence should instead be computed on the parameters of the first layer, the word embedding layer in most \acrshort{llm}s. Early layers are presumed to encode features that are more faithful to the specific tokens and structure of an individual input, rather than the highly abstract concepts found in deeper layers. This specificity makes them less susceptible to the cancellation effect. While the word embedding layer itself may not capture high-level semantic information, its \emph{gradients} do. Because gradients are back-propagated through the entire network via the chain rule, the gradients with respect to the embedding weights implicitly incorporate information from all subsequent layers.
\\\\
Empirical evaluations in \cite{DBLP:conf/nips/YehTSLR22} support this "first is better than last" hypothesis. Using a case deletion metric, they demonstrate that an influence method operating on the word embedding layer significantly outperforms equivalent methods applied to the final layer. Their analysis reveals a clear trend: the quality of the influence score degrades as it is computed on progressively later layers of the model, with the first layer providing the most reliable explanations. The authors provide a strong, empirically grounded rationale for selecting a specific architectural component, the embedding layer, as a cornerstone for constructing efficient and effective instance-based explanations. With this in mind, the following thesis will also shed light onto different layer depths and their effect on instance- and gradient-based explanations.